---
title: "Assignment 6: GLMs (Linear Regressions, ANOVA, & t-tests)"
author: "Laura Martinez"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed

#install.packages('formatR')
library(formatR)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=60), tidy=TRUE)
```

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A06_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup}
#1
setwd("~/Documents/EDA-Fall2022")
getwd()

library(tidyverse)
library(agricolae)
library(lubridate)

LTER_Chem <- read.csv("~/Documents/EDA-Fall2022/Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", stringsAsFactors = TRUE)

LTER_Chem$sampledate <- as.Date(LTER_Chem$sampledate, format = "%m/%d/%y")

#2
Laurastheme <- theme_classic(base_size = 12) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "right")
theme_set(Laurastheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: The mean lake temperatures recorded in July do not change with depth, the difference of means = 0. 
Ha: The mean lake temperatures recorded in July change with depth, the difference of means is != 0. 

4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
LTER_July <-
  LTER_Chem %>%
  filter(daynum %in% c(183:213)) %>%
  select(lakename:daynum, depth:temperature_C) %>%
  drop_na()

#5
plot_tempbydepth <-
  ggplot(LTER_July, aes(x = depth, y = temperature_C)) +
  geom_point() +
  geom_smooth(method = 'lm', se = FALSE, color = "Blue") +
  ylim(0,35) +
  ylab("Temperature (C)") +
  xlab("Depth (meters)") 
print(plot_tempbydepth) 

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The distribution of points suggests that there is a negative relationship between depth and temperature. As depth increases, temperature values decrease. This pattern is observable in ocean and water bodies where temperature decreases as we dive deeper. Furthermore, this trend is shown by the negative linear regression plotted.


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7

July.regression <- lm(LTER_July$temperature_C ~ LTER_July$depth)
print(July.regression)
summary(July.regression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: The overall R-squared is about 73.71%, which means that 73.71% of the values for temperature can be explained due to the depth variable. This shows a strong relationship between depth and temperature. The degrees of freedom is 9671, which is the number of values that can vary in the dataset minus 1. The linear regression model estimates that for every 1m, temperature will decrease by 0.38 degrees Celcius. 

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
July.AIC <- lm(data = LTER_July, temperature_C ~ year4 + daynum + depth)
step(July.AIC)
summary(July.AIC)

#10
AIC(July.regression, July.AIC)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The overall AIC decreases when including more variables. The first model that only had depth had an AIC of 53544.72. By including additional variables such as year4 and daynum, the overall AIC drops down to 53451. This is a reduction of 93.72, and lower AIC values are indicative of models that better explain the variability of the dependent varaible due to the indepenedent variables. In this case, the temperature is better explained by a model that has more varaibles: depth, daynum, and year4. 

---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12

Lakes.anova <- aov(data = LTER_July, temperature_C ~ lakename)
summary(Lakes.anova)

# Format ANOVA as lm
Lakes.anova2 <- lm(data = LTER_July, temperature_C ~ lakename)
summary(Lakes.anova2)

Lakes.groups <- HSD.test(Lakes.anova, "lakename", group = TRUE)
Lakes.groups

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: There is a significant difference in mean temperatures among the lakes. This is especially true when analyzing the groups. Based on the mean temperatures, there are 5 groups that were divided up and share similar means. 


14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
plot_lakes <-
  ggplot(LTER_July, aes(x = depth, y = temperature_C)) +
  geom_point(aes(color = lakename), alpha = 0.5) +
  geom_smooth(method = 'lm', se = FALSE, aes(color = lakename)) +
  ylim(0,35) +
  ylab("Temperature (C)") +
  xlab("Depth (meters)") 
print(plot_lakes) 

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(Lakes.anova)

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Peter Lake has the same mean temperature as Ward Lake and Paul Lake. This is shown by their grouping in category "c" and by Tukey's test that shows a difference that is less than 0.5 for Paul Lake and 1.1 for Ward Lake when compared. There is not a lake that has a statistically distinct mean from all other lakes. This can be seen in the lake groupings which show that all lakes are grouped with at least one other lake. 


17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: Antoher test we can perfrom is the Bartlett's test, which tests if the variances in each of the groups (in this case Peter and Paul lake) are the same.


18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
LTER.subset <- filter(LTER_July, lakename %in% c("Crampton Lake", "Ward Lake"))


July.twosample <- t.test(LTER.subset$temperature_C ~ LTER.subset$lakename)
July.twosample
```

>Answer: Based on the T-test, Crampton Lake has a mean temperature of 15.48 and Ward Lake has a mean temperature of 14.45. These resulst are the same reported values as the aov. The mean temperatures of both lakes are statistically the same because they both are grouped under  "b" and have a difference of -1.02269799. 
