---
title: "Assignment 09: Data Scraping"
author: "Laura Martinez"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A09_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Check your working directory
* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Set your ggplot theme

```{r, message = FALSE}
#1
setwd("~/Documents/EDA-Fall2022")
getwd()

#install.packages(tidyverse)
library(tidyverse)
#install.packages(rvest)
library(rvest)
#install.packages(lubridate)
library(lubridate)

#Set theme
Laurastheme <- theme_classic(base_size = 12) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "right")
theme_set(Laurastheme)

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2021 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2021>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2
theURL <- 'https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2021'
the_main_URL <- read_html(theURL)
```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PSWID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Daily Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings), with the first value being "27.6400".

```{r scrape.the.data}
#3
water.system.name <- the_main_URL %>% html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% html_text()
  
pswid <- the_main_URL %>% html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% html_text()
  
ownership <- the_main_URL %>% html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% html_text()
  
max.withdrawals.mgd <- the_main_URL %>% html_nodes("th~ td+ td") %>% html_text()

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc...

5. Create a line plot of the maximum daily withdrawals across the months for 2021

```{r create.a.dataframe.from.scraped.data}
#4
df <- data.frame("Month" = c(1, 5, 9, 2, 6, 8, 3, 7, 11, 4, 8, 12),
                 "Year" = rep(2021,12),
                 "Water System Name" = water.system.name,
                 "PSWID" = pswid, 
                 "Ownership" = ownership, 
                 "Max_Withdrawals_mgd" = as.numeric(max.withdrawals.mgd)) %>%
  mutate(Date = my(paste(Month,"-",Year)))

#5
max.plot <- ggplot(data = df, aes(x = Date, y = Max_Withdrawals_mgd)) +
  geom_line() +
  xlab("Month") +
  ylab("Maximum Daily Withdrawal")
max.plot

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6.
scrape.it <- function(the_pwsid, the_year){
  
  the_scrape_url <- read_html(paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?', 
                                  'pwsid=', the_pwsid, '&', 'year=', the_year))
  
  #Scrape the data items
  water.system.name <- the_website %>% html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% html_text()
  pswid <- the_website %>% html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% html_text()
  ownership <- the_website %>% html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% html_text()
  max.withdrawals.mgd <- the_website %>% html_nodes("th~ td+ td") %>% html_text()
  
  #Construct dataframe
  dataframe <- data.frame("Month" = c(1, 5, 9, 2, 6, 8, 3, 7, 11, 4, 8, 12),
                   "Year" = rep(the_year,12),
                   "Water System Name" = water.system.name,
                   "PSWID" = pswid, 
                   "Ownership" = ownership, 
                   "Max_Withdrawals_mgd" = as.numeric(max.withdrawals.mgd)) %>%
    mutate(Date = my(paste(Month,"-",Year)))
  return(dataframe)
}
```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7
df_Dur2015 <- scrape.it('03-32-010', 2015)

max.plot_Dur2015 <- ggplot(data = df_Dur2015, aes(x = Date, y = Max_Withdrawals_mgd)) +
  geom_line() +
  xlab("Month") +
  ylab("Maximum Daily Withdrawal")
max.plot_Dur2015

```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8
df_Ash2015 <- scrape.it('01-11-010', 2015)

max.plot_Ash2015 <- ggplot(data = df_Ash2015, aes(x = Date, y = Max_Withdrawals_mgd)) +
  geom_line() +
  xlab("Month") +
  ylab("Maximum Daily Withdrawal")
max.plot_Ash2015

```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2019.Add a smoothed line to the plot.

>TIP: See Section 3.2 in the "09_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data}
#9
ndjsbfjbhdhjsf

```

>Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?
> 